---
title: "Map Rotation"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map Rotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

```{r setup}
library(ggbasemap)
library(ggplot2)
library(sf)

# Load sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

## Introduction

Map rotation allows you to present your spatial data at an angle, which can be useful for:

- Fitting rectangular maps better on page layouts
- Emphasizing certain geographic features
- Creating visually interesting presentations
- Aligning with natural geographic orientations (e.g., coastlines, mountain ranges)

**ggbasemap** uses the **Oblique Mercator** projection to achieve smooth rotation while maintaining accurate geographic relationships.

## Basic Rotation

### Simple Example

To rotate a map, use the `coord_rotate()` function:

```{r basic-rotation, eval=FALSE}
# Get rotation parameters for 45 degrees
rot <- coord_rotate(nc, angle = 45)

# Create the rotated map
ggplot() +
  add_basemap(nc, bbox = rot$bbox) +
  geom_sf(data = nc) +
  rot$coord +
  ggtitle("North Carolina rotated 45°")
```

### What coord_rotate() Returns

The `coord_rotate()` function returns a list with:

- **`coord`**: A `coord_sf` object that applies the rotation
- **`bbox`**: An expanded bounding box that covers the rotated area
- **`crs`**: The coordinate reference system string used
- **`angle`**: The rotation angle in degrees

## Understanding the Components

### The Expanded Bounding Box

When you rotate a map, the corners extend beyond the original bounding box. The `bbox` returned by `coord_rotate()` is expanded to ensure the entire rotated map is covered:

```{r expanded-bbox, eval=FALSE}
# Original bbox
orig_bbox <- sf::st_bbox(nc)
print(orig_bbox)

# Rotated bbox (notice it's larger)
rot <- coord_rotate(nc, 45)
print(rot$bbox)
```

This expanded bbox is used by `add_basemap()` to fetch enough tiles to cover the entire rotated view.

### The Coordinate System

The rotation uses the **Oblique Mercator** projection with these parameters:

- **lat_0**: Latitude of the centroid
- **lonc**: Longitude of the centroid  
- **gamma**: Rotation angle in degrees
- **alpha**: Zero (no skew)
- **k**: Scale factor (1.0)

This projection preserves:
- ✅ Local shapes and angles
- ✅ Straight lines (great circles)
- ✅ Scale along the central line

## Rotation Angles

### Common Angles

```{r angles, eval=FALSE}
# No rotation (0°)
rot0 <- coord_rotate(nc, 0)
ggplot() +
  add_basemap(nc, bbox = rot0$bbox) +
  geom_sf(data = nc) +
  rot0$coord +
  ggtitle("0° (No Rotation)")

# 30°
rot30 <- coord_rotate(nc, 30)
ggplot() +
  add_basemap(nc, bbox = rot30$bbox) +
  geom_sf(data = nc) +
  rot30$coord +
  ggtitle("30° Rotation")

# 45°
rot45 <- coord_rotate(nc, 45)
ggplot() +
  add_basemap(nc, bbox = rot45$bbox) +
  geom_sf(data = nc) +
  rot45$coord +
  ggtitle("45° Rotation")

# 60°
rot60 <- coord_rotate(nc, 60)
ggplot() +
  add_basemap(nc, bbox = rot60$bbox) +
  geom_sf(data = nc) +
  rot60$coord +
  ggtitle("60° Rotation")

# 90°
rot90 <- coord_rotate(nc, 90)
ggplot() +
  add_basemap(nc, bbox = rot90$bbox) +
  geom_sf(data = nc) +
  rot90$coord +
  ggtitle("90° Rotation")
```

### Negative Angles

You can also use negative angles for clockwise rotation:

```{r negative, eval=FALSE}
# -45° (clockwise)
rot_neg45 <- coord_rotate(nc, -45)
ggplot() +
  add_basemap(nc, bbox = rot_neg45$bbox) +
  geom_sf(data = nc) +
  rot_neg45$coord +
  ggtitle("-45° (Clockwise)")
```

### Angles > 360

Angles greater than 360 are automatically normalized:

```{r large-angles, eval=FALSE}
# 405° = 45°
rot405 <- coord_rotate(nc, 405)  # Same as 45°
```

## Practical Examples

### Example 1: Fitting a Page Layout

Sometimes your data is oriented at an angle that doesn't fit well in a standard rectangular plot:

```{r fit-page, eval=FALSE}
# Create elongated data
elongated_data <- nc

# Original (wastes space)
ggplot() +
  add_basemap(elongated_data) +
  geom_sf(data = elongated_data) +
  ggtitle("Original (wasted space on sides)")

# Rotated (better fit)
rot <- coord_rotate(elongated_data, 30)
ggplot() +
  add_basemap(elongated_data, bbox = rot$bbox) +
  geom_sf(data = elongated_data) +
  rot$coord +
  ggtitle("Rotated 30° (better fit)")
```

### Example 2: Aligning with Coastlines

For coastal areas, rotating to align with the coastline can be visually effective:

```{r coastline, eval=FALSE}
# Hypothetical coastal data
rot <- coord_rotate(coastal_data, -15)  # Align with coastline
ggplot() +
  add_basemap(coastal_data, bbox = rot$bbox) +
  geom_sf(data = coastal_data) +
  rot$coord +
  ggtitle("Aligned with Coastline")
```

### Example 3: Emphasizing a Road or River

Rotate to align with linear features:

```{r linear, eval=FALSE}
# Align with a major highway
rot <- coord_rotate(region_data, 25)
ggplot() +
  add_basemap(region_data, bbox = rot$bbox) +
  geom_sf(data = major_road, color = "red", size = 2) +
  geom_sf(data = region_data, fill = NA) +
  rot$coord
```

## Advanced Rotation

### Combining with Transformations

Rotation works perfectly with image transformations:

```{r with-transforms, eval=FALSE}
rot <- coord_rotate(nc, 45)

ggplot() +
  add_basemap(nc, 
              bbox = rot$bbox,
              grayscale = TRUE,
              contrast = 1.2) +
  geom_sf(data = nc, fill = NA, color = "red") +
  rot$coord +
  ggtitle("Rotated + Styled")
```

### Multiple Layers

You can add multiple layers on top of the rotated basemap:

```{r layers, eval=FALSE}
rot <- coord_rotate(nc, 30)

ggplot() +
  # Base: rotated basemap
  add_basemap(nc, bbox = rot$bbox, alpha = 0.7) +
  # Layer 1: county boundaries
  geom_sf(data = nc, aes(fill = area), alpha = 0.5) +
  # Layer 2: major roads
  geom_sf(data = roads, color = "red", size = 1) +
  # Layer 3: cities
  geom_sf(data = cities, aes(size = population), color = "blue") +
  # Apply rotation to all
  rot$coord +
  scale_fill_viridis_c() +
  theme_void()
```

### Custom Styling

```{r custom-style, eval=FALSE}
rot <- coord_rotate(nc, 45)

ggplot() +
  add_basemap(nc, 
              bbox = rot$bbox,
              zoom = 12,
              saturation = 0.5) +
  geom_sf(data = nc, 
          fill = NA, 
          color = "white",
          linewidth = 1) +
  rot$coord +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid = element_blank()
  ) +
  ggtitle("Night Mode Style",
          subtitle = "45° rotation") +
  theme(
    plot.title = element_text(color = "white", size = 20),
    plot.subtitle = element_text(color = "gray70", size = 14)
  )
```

## The Rotation Matrix

The package also provides a `rot()` function that creates a 2D rotation matrix:

```{r rot-matrix, eval=FALSE}
# Create rotation matrix for 45 degrees
m <- rot(45)
print(m)

# Apply to points
points <- matrix(c(1, 0, 0, 1), ncol = 2)
rotated_points <- points %*% m
```

This is used internally but can be useful for custom calculations.

## Performance Considerations

### Tile Fetching

Rotated maps require more tiles because the bounding box is expanded:

- **0° rotation**: Minimum tiles
- **45° rotation**: ~1.4x more tiles (diagonal expansion)
- **Any rotation**: More tiles than no rotation

This means:
- Rotated maps load slightly slower
- Use reasonable zoom levels
- Cache results when possible

### Memory Usage

The expanded bounding box means:
- Larger raster objects
- More memory usage
- Slower rendering

For large areas or high zoom levels, consider:
- Using lower zoom (11-13 instead of 15+)
- Cropping data before rotation
- Using `padding = 0.05` instead of default 0.1

## Troubleshooting

### Issue: Map Looks Cropped

**Problem**: Parts of your data are cut off

**Solution**: The expanded bbox should prevent this, but if not:

```{r fix-cropped, eval=FALSE}
# Increase padding manually
rot <- coord_rotate(nc, 45)
# Expand bbox further
rot$bbox <- rot$bbox * 1.1  # Add 10% more

ggplot() +
  add_basemap(nc, bbox = rot$bbox) +
  geom_sf(data = nc) +
  rot$coord
```

### Issue: Tiles Not Loading

**Problem**: Rotation works but tiles don't appear

**Solution**: Check that you're using the bbox from `coord_rotate()`:

```{r fix-tiles, eval=FALSE}
# Wrong - uses original bbox
ggplot() +
  add_basemap(nc) +  # Missing bbox!
  geom_sf(data = nc) +
  rot$coord

# Correct - uses expanded bbox
ggplot() +
  add_basemap(nc, bbox = rot$bbox) +  # Correct!
  geom_sf(data = nc) +
  rot$coord
```

### Issue: Text Labels Rotate Too

**Problem**: Text labels on your map rotate with the map

**Solution**: This is expected! Everything rotates together. If you need horizontal text, add it after applying the coordinate system with `annotate()` or use `geom_text()` with `hjust` and `vjust` adjustments.

## Best Practices

### 1. Choose the Right Angle

- **0-15°**: Subtle, good for minor adjustments
- **15-45°**: Moderate, good for fitting page layouts
- **45-60°**: Strong rotation, dramatic effect
- **60-90°**: Very strong, use sparingly

### 2. Test Multiple Angles

```{r test-angles, eval=FALSE}
# Try several angles to find the best
angles <- c(0, 15, 30, 45)
plots <- lapply(angles, function(a) {
  rot <- coord_rotate(nc, a)
  ggplot() +
    add_basemap(nc, bbox = rot$bbox) +
    geom_sf(data = nc) +
    rot$coord +
    ggtitle(paste(a, "degrees"))
})

# View all with patchwork or gridExtra
```

### 3. Consider Your Audience

- **Academic papers**: Subtle rotation (0-20°)
- **Presentations**: Moderate rotation (20-45°)
- **Marketing/Design**: Any angle that works

### 4. Combine with Other Features

Rotation works great with:
- Image transformations (grayscale, contrast)
- Different tile sources
- Multiple data layers
- Custom themes

## Complete Example

Here's a complete, publication-ready example:

```{r complete-example, eval=FALSE}
library(ggbasemap)
library(ggplot2)
library(sf)
library(viridis)

# Load data
nc <- st_read(system.file("shape/nc.shp", package = "sf"))

# Calculate rotation (30 degrees looks good for this shape)
rot <- coord_rotate(nc, 30)

# Create the map
map <- ggplot() +
  # Base layer: styled basemap
  add_basemap(nc, 
              bbox = rot$bbox,
              zoom = 9,
              saturation = 0.4,
              brightness = 1.05,
              contrast = 1.1) +
  # Data layer: counties with fill
  geom_sf(data = nc, 
          aes(fill = AREA),
          color = "white",
          linewidth = 0.3,
          alpha = 0.8) +
  # Apply rotation
  rot$coord +
  # Color scale
  scale_fill_viridis_c(option = "magma",
                       name = "Area (sq km)",
                       labels = scales::comma) +
  # Theme
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
  ) +
  # Labels
  labs(title = "North Carolina Counties",
       subtitle = "Rotated 30° for optimal layout",
       caption = "Data: North Carolina county boundaries")

# Save
#ggsave("nc_rotated_map.png", map, width = 10, height = 8, dpi = 300)

print(map)
```

## Next Steps

- Read about **[Image Transformations](transformations.html)** to style your basemaps
- Check **[Advanced Usage](advanced-usage.html)** for power user features
- See **[FAQ](faq.html)** for common questions

## References

- Snyder, J. P. (1987). *Map Projections: A Working Manual*
- OpenStreetMap Wiki: [Mercator Projection](https://wiki.openstreetmap.org/wiki/Mercator)
- PROJ documentation: [Oblique Mercator](https://proj.org/operations/projections/omerc.html)
