---
title: "Map Rotation"
author: "Cédric Bouffard"
date: "2026-03-14"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map Rotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  eval = TRUE
)
```

```{r setup}
library(ggbasemap)
library(ggplot2)
library(sf)

# Load sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

## Introduction

Map rotation allows you to present your spatial data at an angle, which can be useful for:

- Fitting rectangular maps better on page layouts
- Emphasizing certain geographic features
- Creating visually interesting presentations
- Aligning with natural geographic orientations (e.g., coastlines, mountain ranges)

**ggbasemap** uses the **Oblique Mercator** projection to achieve smooth rotation while maintaining accurate geographic relationships.

## Basic Rotation

To rotate a map, use the `coord_rotate()` function:

```{r basic-rotation, fig.cap="Map rotated 45 degrees"}
# Get rotation parameters for 45 degrees
rot <- coord_rotate(nc, angle = 45)

# Create the rotated map
ggplot() +
  add_basemap(nc, bbox = rot$bbox) +
  geom_sf(data = nc) +
  rot$coord +
  ggtitle("North Carolina rotated 45°")
```

The `coord_rotate()` function returns a list with:

- **`coord`**: A `coord_sf` object that applies the rotation
- **`bbox`**: An expanded bounding box that covers the rotated area
- **`crs`**: The coordinate reference system string used
- **`angle`**: The rotation angle in degrees

## Rotation Angles

Different rotation angles produce different visual effects:

```{r angle-comparison, fig.cap="Comparison of rotation angles"}
# Create a 2-panel comparison
library(patchwork)

# No rotation
rot0 <- coord_rotate(nc, 0)
p0 <- ggplot() +
  add_basemap(nc, bbox = rot0$bbox) +
  geom_sf(data = nc, fill = NA, color = "red") +
  rot0$coord +
  ggtitle("0° (No Rotation)") +
  theme_void()

# 45 degree rotation
rot45 <- coord_rotate(nc, 45)
p45 <- ggplot() +
  add_basemap(nc, bbox = rot45$bbox) +
  geom_sf(data = nc, fill = NA, color = "blue") +
  rot45$coord +
  ggtitle("45° Rotation") +
  theme_void()

# Display side by side
p0 + p45
```

## Combining with Transformations

Rotation works perfectly with image transformations:

```{r with-transforms, fig.cap="Rotated with styled basemap"}
rot <- coord_rotate(nc, 45)

ggplot() +
  add_basemap(nc, 
              bbox = rot$bbox,
              grayscale = TRUE,
              contrast = 1.2) +
  geom_sf(data = nc, fill = NA, color = "red") +
  rot$coord +
  ggtitle("Rotated + Styled")
```

## Controlling Aspect Ratio

You can control the aspect ratio (width/height) of the rotated map using the `ratio` parameter:

```{r ratio-example, fig.cap="Different aspect ratios"}
# Create comparison of different ratios
rot_169 <- coord_rotate(nc, 45, ratio = 16/9)
rot_11 <- coord_rotate(nc, 45, ratio = 1)

p169 <- ggplot() +
  add_basemap(nc, bbox = rot_169$bbox) +
  geom_sf(data = nc, fill = NA, color = "darkblue") +
  rot_169$coord +
  ggtitle("16:9 Widescreen") +
  theme_void()

p11 <- ggplot() +
  add_basemap(nc, bbox = rot_11$bbox) +
  geom_sf(data = nc, fill = NA, color = "darkgreen") +
  rot_11$coord +
  ggtitle("1:1 Square") +
  theme_void()

p169 + p11
```

Common ratio values:
- `16/9` for widescreen displays and presentations
- `4/3` for standard presentation format
- `1` for square format (social media, posters)
- Custom values like `2` or `0.5` for specific layouts

## Complete Example

Here's a complete example combining multiple features:

```{r complete-example, fig.cap="Complete example with rotation"}
library(viridis)

# Calculate rotation
rot <- coord_rotate(nc, 30)

# Create the map
ggplot() +
  add_basemap(nc, 
              bbox = rot$bbox,
              saturation = 0.4,
              brightness = 1.05) +
  geom_sf(data = nc, 
          aes(fill = AREA),
          color = "white",
          linewidth = 0.3,
          alpha = 0.8) +
  rot$coord +
  scale_fill_viridis_c(option = "magma") +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "North Carolina Counties",
       subtitle = "Rotated 30° for optimal layout")
```

## Next Steps

- Read about **[Image Transformations](transformations.html)** to style your basemaps
- Check the **FAQ** for common questions