---
title: "Map Rotation"
author: "Cédric Bouffard"
date: "2026-03-14"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map Rotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  eval = TRUE
)
```

```{r setup}
library(ggbasemap)
library(ggplot2)
library(sf)

# Load sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

## Introduction

Map rotation allows you to present your spatial data at an angle, which can be useful for:

- Fitting rectangular maps better on page layouts
- Emphasizing certain geographic features
- Creating visually interesting presentations
- Aligning with natural geographic orientations (e.g., coastlines, mountain ranges)

**ggbasemap** uses the **Oblique Mercator** projection to achieve smooth rotation while maintaining accurate geographic relationships.

## Basic Rotation

### Simple Example

To rotate a map, use the `coord_rotate()` function:

```{r basic-rotation, fig.cap="Map rotated 45 degrees"}
# Get rotation parameters for 45 degrees
rot <- coord_rotate(nc, angle = 45)

# Create the rotated map
ggplot() +
  add_basemap(nc, bbox = rot$bbox) +
  geom_sf(data = nc) +
  rot$coord +
  ggtitle("North Carolina rotated 45°")
```

### What coord_rotate() Returns

The `coord_rotate()` function returns a list with:

- **`coord`**: A `coord_sf` object that applies the rotation
- **`bbox`**: An expanded bounding box that covers the rotated area
- **`crs`**: The coordinate reference system string used
- **`angle`**: The rotation angle in degrees

## Understanding the Components

### The Expanded Bounding Box

When you rotate a map, the corners extend beyond the original bounding box. The `bbox` returned by `coord_rotate()` is expanded to ensure the entire rotated map is covered:

```{r expanded-bbox}
# Original bbox
orig_bbox <- sf::st_bbox(nc)
print(orig_bbox)

# Rotated bbox (notice it's larger)
rot <- coord_rotate(nc, 45)
print(rot$bbox)
```

## Rotation Angles

### Common Angles

```{r angle-0, fig.cap="No rotation (0°)"}
rot0 <- coord_rotate(nc, 0)
ggplot() +
  add_basemap(nc, bbox = rot0$bbox) +
  geom_sf(data = nc) +
  rot0$coord +
  ggtitle("0° (No Rotation)")
```

```{r angle-30, fig.cap="30° rotation"}
rot30 <- coord_rotate(nc, 30)
ggplot() +
  add_basemap(nc, bbox = rot30$bbox) +
  geom_sf(data = nc) +
  rot30$coord +
  ggtitle("30° Rotation")
```

```{r angle-45, fig.cap="45° rotation"}
rot45 <- coord_rotate(nc, 45)
ggplot() +
  add_basemap(nc, bbox = rot45$bbox) +
  geom_sf(data = nc) +
  rot45$coord +
  ggtitle("45° Rotation")
```

## Practical Examples

### Example 1: Fitting a Page Layout

Sometimes your data is oriented at an angle that doesn't fit well in a standard rectangular plot:

```{r fit-page, fig.cap="Rotated for better page fit"}
# Rotated (better fit)
rot <- coord_rotate(nc, 30)
ggplot() +
  add_basemap(nc, bbox = rot$bbox) +
  geom_sf(data = nc) +
  rot$coord +
  ggtitle("Rotated 30° (better fit)")
```

## Advanced Rotation

### Combining with Transformations

Rotation works perfectly with image transformations:

```{r with-transforms, fig.cap="Rotated with styled basemap"}
rot <- coord_rotate(nc, 45)

ggplot() +
  add_basemap(nc, 
              bbox = rot$bbox,
              grayscale = TRUE,
              contrast = 1.2) +
  geom_sf(data = nc, fill = NA, color = "red") +
  rot$coord +
  ggtitle("Rotated + Styled")
```

### Multiple Layers

You can add multiple layers on top of the rotated basemap:

```{r layers, fig.cap="Multiple layers with rotation"}
rot <- coord_rotate(nc, 30)

ggplot() +
  # Base: rotated basemap
  add_basemap(nc, bbox = rot$bbox, alpha = 0.7) +
  # Layer 1: county boundaries
  geom_sf(data = nc, aes(fill = AREA), alpha = 0.5) +
  # Apply rotation to all
  rot$coord +
  scale_fill_viridis_c() +
  theme_void()
```

### Custom Styling

```{r custom-style, fig.cap="Night mode style with rotation"}
rot <- coord_rotate(nc, 45)

ggplot() +
  add_basemap(nc, 
              bbox = rot$bbox,
              zoom = 12,
              saturation = 0.5) +
  geom_sf(data = nc, 
          fill = NA, 
          color = "white",
          linewidth = 1) +
  rot$coord +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    panel.grid = element_blank()
  ) +
  ggtitle("Night Mode Style",
          subtitle = "45° rotation") +
  theme(
    plot.title = element_text(color = "white", size = 20),
    plot.subtitle = element_text(color = "gray70", size = 14)
  )
```

## Complete Example

## Controlling Aspect Ratio

You can control the aspect ratio (width/height) of the rotated map using the `ratio` parameter. This is useful for fitting maps to specific output formats:

### Common Aspect Ratios

```{r ratio-169, fig.cap="Widescreen 16:9 ratio"}
# 16:9 - Perfect for presentations and widescreen displays
rot_16_9 <- coord_rotate(nc, 45, ratio = 16/9)

ggplot() +
  add_basemap(nc, bbox = rot_16_9$bbox) +
  geom_sf(data = nc, fill = NA, color = "red", linewidth = 1) +
  rot_16_9$coord +
  ggtitle("16:9 Widescreen Format")
```

```{r ratio-43, fig.cap="Standard 4:3 ratio"}
# 4:3 - Classic presentation format
rot_4_3 <- coord_rotate(nc, 45, ratio = 4/3)

ggplot() +
  add_basemap(nc, bbox = rot_4_3$bbox) +
  geom_sf(data = nc, fill = NA, color = "blue", linewidth = 1) +
  rot_4_3$coord +
  ggtitle("4:3 Standard Format")
```

```{r ratio-square, fig.cap="Square 1:1 ratio"}
# 1:1 - Square format (good for Instagram, posters)
rot_square <- coord_rotate(nc, 45, ratio = 1)

ggplot() +
  add_basemap(nc, bbox = rot_square$bbox) +
  geom_sf(data = nc, fill = NA, color = "green", linewidth = 1) +
  rot_square$coord +
  ggtitle("1:1 Square Format")
```

### How It Works

When you specify a ratio:
- The function calculates the natural bbox of your rotated data
- If the natural ratio is different from your target ratio, it expands the bbox to match
- The bbox is always expanded (never cropped) to ensure all data is visible
- Both the tile bbox and the coordinate limits are adjusted

### Practical Example: Publication Figure

```{r ratio-publication, fig.cap="Publication-ready figure with custom ratio"}
# Create a 2:1 ratio figure for a journal article
rot_pub <- coord_rotate(nc, 30, ratio = 2)

ggplot() +
  add_basemap(nc, 
              bbox = rot_pub$bbox,
              saturation = 0.5,
              contrast = 1.1) +
  geom_sf(data = nc, 
          aes(fill = AREA),
          color = "white",
          linewidth = 0.3) +
  rot_pub$coord +
  scale_fill_viridis_c() +
  theme_void() +
  theme(legend.position = "bottom") +
  ggtitle("North Carolina - 2:1 Aspect Ratio",
          subtitle = "Perfect for journal figures")
```

## Complete Example

Here's a complete, publication-ready example:

```{r complete-example, fig.cap="Complete example with rotation"}
library(viridis)

# Calculate rotation (30 degrees looks good for this shape)
rot <- coord_rotate(nc, 30)

# Create the map
ggplot() +
  # Base layer: styled basemap
  add_basemap(nc, 
              bbox = rot$bbox,
              zoom = 9,
              saturation = 0.4,
              brightness = 1.05,
              contrast = 1.1) +
  # Data layer: counties with fill
  geom_sf(data = nc, 
          aes(fill = AREA),
          color = "white",
          linewidth = 0.3,
          alpha = 0.8) +
  # Apply rotation
  rot$coord +
  # Color scale
  scale_fill_viridis_c(option = "magma",
                       name = "Area (sq km)",
                       labels = scales::comma) +
  # Theme
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
  ) +
  # Labels
  labs(title = "North Carolina Counties",
       subtitle = "Rotated 30° for optimal layout",
       caption = "Data: North Carolina county boundaries")
```

## Next Steps

- Read about **[Image Transformations](transformations.html)** to style your basemaps
- Check the **FAQ** for common questions
