---
title: "Image Transformations"
author: "Cédric Bouffard"
date: "2026-03-14"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Image Transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  eval = TRUE
)
```

```{r setup}
library(ggbasemap)
library(ggplot2)
library(sf)

# Load sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

## Introduction

**ggbasemap** provides powerful image transformation capabilities that allow you to style your basemaps to match your visualization needs. You can adjust grayscale, saturation, brightness, contrast, and gamma correction.

## Available Transformations

### 1. Grayscale (Black & White)

Convert your basemap to black and white for a classic look:

```{r grayscale, fig.cap="Grayscale basemap"}
ggplot() +
  add_basemap(nc, grayscale = TRUE) +
  geom_sf(data = nc, fill = NA, color = "white", linewidth = 1) +
  theme_void()
```

This is perfect for:
- Academic publications
- Print media
- When you want data to stand out
- Creating vintage-style maps

### 2. Saturation

Control the intensity of colors:

```{r saturation-normal, fig.cap="Normal saturation"}
ggplot() +
  add_basemap(nc, saturation = 1) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Saturation = 1 (default)")
```

```{r saturation-low, fig.cap="Reduced saturation"}
ggplot() +
  add_basemap(nc, saturation = 0.5) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Saturation = 0.5")
```

### 3. Brightness

Adjust the overall lightness:

```{r brightness, fig.cap="Adjusted brightness"}
# Darker (-30%)
ggplot() +
  add_basemap(nc, brightness = 0.7) +
  geom_sf(data = nc, fill = NA, color = "yellow") +
  ggtitle("Brightness = 0.7 (darker)")
```

### 4. Contrast

Enhance or reduce the difference between light and dark areas:

```{r contrast, fig.cap="High contrast"}
ggplot() +
  add_basemap(nc, contrast = 1.5) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Contrast = 1.5 (punchy)")
```

### 5. Gamma Correction

Adjust midtones without affecting blacks and whites:

```{r gamma, fig.cap="Gamma correction"}
ggplot() +
  add_basemap(nc, gamma = 0.8) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Gamma = 0.8 (brighter midtones)")
```

## Combining Transformations

The real power comes from combining multiple transformations:

### Vintage/Sepia Effect

```{r vintage, fig.cap="Vintage style map"}
ggplot() +
  add_basemap(nc,
              saturation = 0.3,
              brightness = 0.85,
              contrast = 1.2,
              gamma = 1.1) +
  geom_sf(data = nc, fill = NA, color = "brown", linewidth = 0.8) +
  theme_void() +
  ggtitle("Vintage Style")
```

### High-Contrast Monochrome

```{r high-contrast, fig.cap="High contrast monochrome"}
ggplot() +
  add_basemap(nc,
              grayscale = TRUE,
              brightness = 1.1,
              contrast = 1.6) +
  geom_sf(data = nc, fill = NA, color = "black") +
  theme_void() +
  ggtitle("High-Contrast Monochrome")
```

### Soft/Faded Look

```{r soft, fig.cap="Soft faded look"}
ggplot() +
  add_basemap(nc,
              saturation = 0.6,
              brightness = 1.1,
              contrast = 0.8,
              gamma = 0.9) +
  geom_sf(data = nc, fill = NA, color = "darkblue") +
  theme_void() +
  ggtitle("Soft/Faded Look")
```

### Data-Friendly Background

When you want your data to be the star:

```{r data-friendly, fig.cap="Data-friendly background"}
ggplot() +
  add_basemap(nc,
              saturation = 0.2,
              brightness = 1.05,
              contrast = 0.9) +
  geom_sf(data = nc, 
          aes(fill = AREA),
          color = "white",
          linewidth = 0.3) +
  scale_fill_viridis_c() +
  theme_void() +
  ggtitle("Data-Friendly Background")
```

## Complete Reference Table

| Parameter | Default | Range | Effect |
|-----------|---------|-------|--------|
| `grayscale` | FALSE | TRUE/FALSE | Convert to B&W |
| `saturation` | 1 | 0 to Inf | Color intensity (0=grayscale) |
| `brightness` | 1 | 0 to Inf | Overall lightness |
| `contrast` | 1 | 0 to Inf | Difference between light/dark |
| `gamma` | 1 | 0 to Inf | Midtones adjustment |

## Transformation Order

Transformations are applied in this order:

1. **Gamma** (non-linear correction)
2. **Brightness** (linear multiplication)
3. **Contrast** (centered on 0.5)
4. **Saturation** (color adjustment)
5. **Grayscale** (final conversion if TRUE)

This order ensures the best visual results.

## Best Practices

### For Print Publications

```{r print, fig.cap="Print-ready grayscale"}
ggplot() +
  add_basemap(nc,
              grayscale = TRUE,
              contrast = 1.2,
              brightness = 0.95) +
  geom_sf(data = nc)
```

### For Web/Digital Display

```{r web, fig.cap="Web-optimized colors"}
ggplot() +
  add_basemap(nc,
              saturation = 1.1,
              brightness = 1.05,
              contrast = 1.0) +
  geom_sf(data = nc)
```

## Common Mistakes to Avoid

1. **Don't overdo it**: Extreme values (>2 or <0.3) usually look bad
2. **Check your data**: Make sure your data colors work with the basemap
3. **Test on different screens**: What looks good on your monitor might not on others

## Next Steps

- Read about **[Map Rotation](rotation.html)** to create angled perspectives
- Check the **FAQ** for common questions
