---
title: "Image Transformations"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Image Transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
```

```{r setup}
library(ggbasemap)
library(ggplot2)
library(sf)

# Load sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

## Introduction

**ggbasemap** provides powerful image transformation capabilities that allow you to style your basemaps to match your visualization needs. You can adjust grayscale, saturation, brightness, contrast, and gamma correction.

## Available Transformations

### 1. Grayscale (Black & White)

Convert your basemap to black and white for a classic look:

```{r grayscale, eval=FALSE}
ggplot() +
  add_basemap(nc, grayscale = TRUE) +
  geom_sf(data = nc, fill = NA, color = "white", linewidth = 1) +
  theme_void()
```

This is perfect for:
- Academic publications
- Print media
- When you want data to stand out
- Creating vintage-style maps

### 2. Saturation

Control the intensity of colors:

```{r saturation, eval=FALSE}
# No saturation (black & white)
ggplot() +
  add_basemap(nc, saturation = 0) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Saturation = 0")

# Reduced saturation (50%)
ggplot() +
  add_basemap(nc, saturation = 0.5) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Saturation = 0.5")

# Normal saturation
ggplot() +
  add_basemap(nc, saturation = 1) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Saturation = 1 (default)")

# Increased saturation (150%)
ggplot() +
  add_basemap(nc, saturation = 1.5) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Saturation = 1.5")
```

**Use cases:**
- `saturation = 0`: Documents, printing
- `saturation = 0.3-0.7`: Subtle, professional look
- `saturation = 1`: Normal (default)
- `saturation = 1.2-2`: Vibrant, eye-catching maps

### 3. Brightness

Adjust the overall lightness:

```{r brightness, eval=FALSE}
# Darker (-30%)
ggplot() +
  add_basemap(nc, brightness = 0.7) +
  geom_sf(data = nc, fill = NA, color = "yellow") +
  ggtitle("Brightness = 0.7 (darker)")

# Normal
ggplot() +
  add_basemap(nc, brightness = 1) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Brightness = 1 (default)")

# Brighter (+30%)
ggplot() +
  add_basemap(nc, brightness = 1.3) +
  geom_sf(data = nc, fill = NA, color = "darkblue") +
  ggtitle("Brightness = 1.3 (brighter)")
```

**Tips:**
- Use `brightness < 1` with light-colored data
- Use `brightness > 1` with dark-colored data
- Combine with `contrast` for best results

### 4. Contrast

Enhance or reduce the difference between light and dark areas:

```{r contrast, eval=FALSE}
# Low contrast (flat)
ggplot() +
  add_basemap(nc, contrast = 0.6) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Contrast = 0.6 (flat)")

# Normal
ggplot() +
  add_basemap(nc, contrast = 1) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Contrast = 1 (default)")

# High contrast (punchy)
ggplot() +
  add_basemap(nc, contrast = 1.5) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Contrast = 1.5 (punchy)")
```

**When to use:**
- `contrast < 1`: Soft, dreamy aesthetics
- `contrast > 1`: Sharp, dramatic effect
- `contrast = 1.2-1.4`: Good for satellite imagery

### 5. Gamma Correction

Adjust midtones without affecting blacks and whites:

```{r gamma, eval=FALSE}
# Gamma < 1: brighter midtones (good for dark imagery)
ggplot() +
  add_basemap(nc, gamma = 0.8) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Gamma = 0.8 (brighter midtones)")

# Gamma = 1: no change
ggplot() +
  add_basemap(nc, gamma = 1) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Gamma = 1 (default)")

# Gamma > 1: darker midtones (good for washed-out imagery)
ggplot() +
  add_basemap(nc, gamma = 1.2) +
  geom_sf(data = nc, fill = NA, color = "red") +
  ggtitle("Gamma = 1.2 (darker midtones)")
```

**Technical note:** Gamma correction is non-linear and applied first, before brightness and contrast.

## Combining Transformations

The real power comes from combining multiple transformations:

### Vintage/Sepia Effect

```{r vintage, eval=FALSE}
ggplot() +
  add_basemap(nc,
              saturation = 0.3,      # Desaturate
              brightness = 0.85,     # Slightly darker
              contrast = 1.2,        # More contrast
              gamma = 1.1) +         # Darker midtones
  geom_sf(data = nc, fill = NA, color = "brown", linewidth = 0.8) +
  theme_void() +
  ggtitle("Vintage Style")
```

### High-Contrast Monochrome

```{r high-contrast, eval=FALSE}
ggplot() +
  add_basemap(nc,
              grayscale = TRUE,      # Black & white
              brightness = 1.1,      # Brighter
              contrast = 1.6) +      # Very contrasty
  geom_sf(data = nc, fill = NA, color = "black") +
  theme_void() +
  ggtitle("High-Contrast Monochrome")
```

### Soft/Faded Look

```{r soft, eval=FALSE}
ggplot() +
  add_basemap(nc,
              saturation = 0.6,      # Muted colors
              brightness = 1.1,      # Brighter
              contrast = 0.8,        # Less contrast
              gamma = 0.9) +         # Brighter midtones
  geom_sf(data = nc, fill = NA, color = "darkblue") +
  theme_void() +
  ggtitle("Soft/Faded Look")
```

### Night Mode

```{r night, eval=FALSE}
ggplot() +
  add_basemap(nc,
              saturation = 0.7,
              brightness = 0.6,      # Much darker
              contrast = 1.3,
              gamma = 1.2) +
  geom_sf(data = nc, fill = NA, color = "cyan") +
  theme_void() +
  ggtitle("Night Mode")
```

### Data-Friendly Background

When you want your data to be the star:

```{r data-friendly, eval=FALSE}
ggplot() +
  add_basemap(nc,
              saturation = 0.2,      # Almost grayscale
              brightness = 1.05,
              contrast = 0.9) +
  geom_sf(data = nc, 
          aes(fill = area),          # Your data stands out
          color = "white",
          linewidth = 0.3) +
  scale_fill_viridis_c() +
  theme_void() +
  ggtitle("Data-Friendly Background")
```

## Complete Reference Table

| Parameter | Default | Range | Effect |
|-----------|---------|-------|--------|
| `grayscale` | FALSE | TRUE/FALSE | Convert to B&W |
| `saturation` | 1 | 0 to Inf | Color intensity (0=grayscale) |
| `brightness` | 1 | 0 to Inf | Overall lightness |
| `contrast` | 1 | 0 to Inf | Difference between light/dark |
| `gamma` | 1 | 0 to Inf | Midtones adjustment |

## Transformation Order

Transformations are applied in this order:

1. **Gamma** (non-linear correction)
2. **Brightness** (linear multiplication)
3. **Contrast** (centered on 0.5)
4. **Saturation** (color adjustment)
5. **Grayscale** (final conversion if TRUE)

This order ensures the best visual results.

## Best Practices

### For Print Publications

```{r print, eval=FALSE}
ggplot() +
  add_basemap(data,
              grayscale = TRUE,      # Always use B&W for print
              contrast = 1.2,        # Boost contrast for printing
              brightness = 0.95) +   # Slightly darker for better ink coverage
  geom_sf(data = data)
```

### For Web/Digital Display

```{r web, eval=FALSE}
ggplot() +
  add_basemap(data,
              saturation = 1.1,      # Slightly more vibrant
              brightness = 1.05,     # Brighter for screens
              contrast = 1.0) +      # Normal contrast
  geom_sf(data = data)
```

### For Satellite Imagery

```{r satellite, eval=FALSE}
ggplot() +
  add_basemap(data,
              url = "...satellite...",
              contrast = 1.3,        # Enhance details
              gamma = 0.9,           # Brighten shadows
              saturation = 1.2) +    # Vivid colors
  geom_sf(data = data)
```

### For Road Maps

```{r roads, eval=FALSE}
ggplot() +
  add_basemap(data,
              saturation = 0.4,      # Muted
              brightness = 1.1,      # Light background
              contrast = 0.9) +      # Softer
  geom_sf(data = data)
```

## Common Mistakes to Avoid

1. **Don't overdo it**: Extreme values (>2 or <0.3) usually look bad
2. **Check your data**: Make sure your data colors work with the basemap
3. **Test on different screens**: What looks good on your monitor might not on others
4. **Save with care**: Transformations affect file size and quality

## Examples Gallery

### Example 1: Minimalist

```{r ex1, eval=FALSE}
ggplot() +
  add_basemap(nc, grayscale = TRUE, brightness = 1.1) +
  geom_sf(data = nc, fill = NA, color = "black") +
  theme_void()
```

### Example 2: Vibrant

```{r ex2, eval=FALSE}
ggplot() +
  add_basemap(nc, saturation = 1.4, contrast = 1.2) +
  geom_sf(data = nc, aes(fill = area), color = "white") +
  scale_fill_viridis_c(option = "plasma")
```

### Example 3: Document-Ready

```{r ex3, eval=FALSE}
ggplot() +
  add_basemap(nc, 
              grayscale = TRUE, 
              contrast = 1.3,
              brightness = 0.9) +
  geom_sf(data = nc, fill = "lightgray", color = "black") +
  theme_bw()
```

## Next Steps

- Learn about **[Map Rotation](rotation.html)** for angled perspectives
- Explore **[Advanced Usage](advanced-usage.html)** for power users
- Check the **[FAQ](faq.html)** for common questions
